Learning features and abstract actions for computing generalized plans

### 2018-09-05 doc/paper.tex
Latex table draft

\def\year{2019}\relax
%File: formatting-instruction.tex
\documentclass[letterpaper]{article} %DO NOT CHANGE THIS
\usepackage{aaai19}  %Required
\usepackage{times}  %Required
\usepackage{helvet}  %Required
\usepackage{courier}  %Required
\usepackage{url}  %Required
\usepackage{graphicx}  %Required
\frenchspacing  %Required
\setlength{\pdfpagewidth}{8.5in}  %Required
\setlength{\pdfpageheight}{11in}  %Required
%PDF Info Is Required:
  \pdfinfo{
/Title ()
/Author ()}
\setcounter{secnumdepth}{0}

@@

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{enumerate}
%\usepackage{upgreek}
%\usepackage[normalem]{ulem}

%\usepackage[small]{caption}

@@

\newcommand{\Omit}[1]{}
\newcommand{\denselist}{\itemsep -1pt\partopsep 0pt}
\newcommand{\tup}[1]{\langle #1 \rangle}
\newcommand{\pair}[1]{\langle #1 \rangle}
% \newcommand{\citeay}[1]{\citeauthor{#1} [\citeyear{#1}]}
\newcommand{\citeay}[1]{\citeauthor{#1} (\citeyear{#1})}
\newcommand{\citeaywithnote}[2]{\citeauthor{#1} (\citeyear[#2]{#1})}

@@

% \pdfinfo{ ..}

@@

Bonet's and Geffner's  formulation of generalized planning, however, assumes that the
features and the abstract actions are given.   In this work, we address this limitation.
We achieving this by  \emph{showing  how features and abstract actions  can be learned from
the primitive predicates used to define the instances and state transitions
sampled from them.} For example,  in \cite{bonet:ijcai2018}, 
the general policy for achieving the goal $clear(x)$  is obtained 
using a FOND planner by introducing the boolean feature $H$ for holding a block,
the numerical feature $n(x)$ for counting the  number of blocks above $x$, and  two   abstract  actions:
one  with  preconditions  $\neg H$ and $n(x) > 0$, and effects $H$ and $n(x)\mminus$
(decrement $n(x)$), the other with precondition $H$ and effect $\neg H$.
In this work, we show how  to obtain such  policies without
having to provide the  features and  abstract actions by hand.
The abstraction is learned instead from the domain predicates
and  state transitions sampled from some  instances.

@@

Bonet's and Geffner's formulation and computational model rest on 
features $F$ and  sound abstract actions $A_F$ provided by hand.
The contribution of this work is to learn them both from samples:

@@

The theory $T({\cal S},{\cal F})$ given by the formulas  \ref{d1})--(\ref{goal}
has a number of atoms that is linear in the size of the feature pool $\cal F$,
and quadratic in the number of expanded  states $s$  and transitions $(s,s')$ in ${\cal S}$.
The resulting number of clauses is ....  , which can be roughly cut in half by  symmetry considerations.

@@

The computation of the feature pool $\cal F$ used in $T({\cal S},{\cal F})$
relies on the predicates used to encode the instances in $\Q$.
A simple grammar   allow  us   to construct new predicates from them, 
and the boolean and numerical features in the pool
are defined from the resulting \emph{unary predicates}. 
The key issue is how to define new predicates from old  ones. For this 
we take a solution off-the-shelf: \emph{description logics (DL)},  called  also
\emph{concept languages},  are grammars for doing exactly this.
Concepts $C$  refer  to predicates of arity 1 and roles $R$  to predicates of arity 2.
Concept languages have been used before for  generating general policies using purely learning methods  \cite{martin:concept,fern:bias}.
For details on the syntax and semantics of description logics; see \cite{dl-handbook}.

@@

  \begin{itemize}
  \item $C \ \leftarrow\ \ C_p, C_u, C_x$, primitive, universal,  nominal c's,\footnote{Denotation of $C_x$ for parameter $x$ is $\{x\}$.}
  \item $C \ \leftarrow\  \neg C, C \sqcap C'$, negation and conjunction, 
  \item $C \ \leftarrow\  \exists R.C , \forall R.C$, concepts that use of  roles: first denotes $x$ s.t. for some  $y$,  $R(x,y) \land C(y)$.
  Second,  $x$ s.t. for all $y$, $R(x,y) \rightarrow C(y)$,
  \item $C \ \leftarrow\  R=R'$, stands for $x$ s.t. set of $y$'s  s.t. $R(x,y)$ equal to set of $y$'s s.t. $R'(x,y)$,
    \item $R \leftarrow R_p, R^{-1}, R^*$ ; primitive, inverses, transitive.
%     \item $R \leftarrow R \circ R'$, role composition: $(x,y)$ in $R \circ R'$ if $(x,z)$ in $R$ and $(z,y)$ in $R'$.
%     \item $R \leftarrow R:C$, denotes pairs $(x,y)$ in $R$ s.t. $C(y)$.
 \end{itemize}

@@

The set of  concepts and roles  with complexity bounded by $k$ is denoted as ${\cal G}(k)$
with their  complexity given by the    minimum number of  rules for generating them.
The size  of this set is reduced by incrementally pruning ``duplicate'' concepts,
those whose denotation coincides with the denotation of a another concept over all
the sampled states in ${\cal S}$.

@@

We evaluate the computational model given by steps 1--6 above over several domains. *Fill up general  details in 1--6
that don't change from problem to problem: how ${\cal S}$ obtained, value of $k$, SAT solver used, FOND planners, translator, etc.
The goal marking scheme that  yields the smaller  theory $T_G({\cal S},{\cal F})$ is obtained an optimal plan over
a few (always less than xxxx) instances $P$.

Experiments setting: cpu, cutoffs, etc.

For each problem: 1--6 details that vary from problem to problem. Follow same structure:
$\Q$, $P$s, primitive predicates, number of atoms and clauses  of $T$ and/or $T^G$,
SAT times,  $F$, $|A_F|$;  number of FOND atoms and actions in $Q^+_F$ (include those in $F$, $A_F$ plus translation), $I_F$ and $G_F$,
FOND time, resulting policy.  Numbers can go all in a table.

>>>>

\def\year{2019}\relax
%File: formatting-instruction.tex
\documentclass[letterpaper]{article} %DO NOT CHANGE THIS
\usepackage{aaai19}  %Required
\usepackage{times}  %Required
\usepackage{helvet}  %Required
\usepackage{courier}  %Required
\usepackage{url}  %Required
\usepackage{graphicx}  %Required
\frenchspacing  %Required
\setlength{\pdfpagewidth}{8.5in}  %Required
\setlength{\pdfpageheight}{11in}  %Required

\setcounter{secnumdepth}{0}

@@

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{enumerate}
%\usepackage{upgreek}
%\usepackage[normalem]{ulem}
\usepackage{multirow}
% \usepackage[small]{caption}

@@

\newcommand{\Omit}[1]{}
\newcommand{\denselist}{\itemsep -1pt\partopsep 0pt}
\newcommand{\tup}[1]{\langle #1 \rangle}
\newcommand{\pair}[1]{\langle #1 \rangle}
\newcommand{\tuple}[1]{\ensuremath{\langle #1 \rangle}}
\newcommand{\set}[1]{\ensuremath{\left\{#1 \right\}}}
\newcommand{\setst}[2]{\ensuremath{\left\{#1 \mid #2 \right\}}}
\newcommand{\abs}[1]{\ensuremath{\left\vert{#1}\right\vert}}
% \newcommand{\citeay}[1]{\citeauthor{#1} [\citeyear{#1}]}
\newcommand{\citeay}[1]{\citeauthor{#1} (\citeyear{#1})}
\newcommand{\citeaywithnote}[2]{\citeauthor{#1} (\citeyear[#2]{#1})}

@@

%PDF Info Is Required (no accents, etc.!):
\pdfinfo{
/Title ()
/Author ()}

@@

Bonet and Geffner's  formulation of generalized planning, however, assumes that the
features and the abstract actions are given.   In this work, we address this limitation.
We achieving this by  \emph{showing  how features and abstract actions  can be learned from
the primitive predicates used to define the instances and state transitions
sampled from them.} For example,  in \cite{bonet:ijcai2018}, 
the general policy for achieving the goal $clear(x)$  is obtained 
using a FOND planner by introducing the boolean feature $H$ for holding a block,
the numerical feature $n(x)$ for counting the  number of blocks above $x$, and  two   abstract  actions:
one  with  preconditions  $\neg H$ and $n(x) > 0$, and effects $H$ and $n(x)\mminus$
(decrement $n(x)$), the other with precondition $H$ and effect $\neg H$.
In this work, we show how  to obtain such  policies without
having to provide the  features and  abstract actions by hand.
The abstraction is learned instead from the domain predicates
and  state transitions sampled from some  instances.

@@

Bonet and Geffner's formulation and computational model rest on 
features $F$ and  sound abstract actions $A_F$ provided by hand.
The contribution of this work is to learn them both from samples:

@@

The theory $T({\cal S},{\cal F})$ given by the formulas  (\ref{d1})--(\ref{goal})
has a number of atoms that is linear in the size of the feature pool $\cal F$,
and quadratic in the number of expanded  states $s$  and transitions $(s,s')$ in ${\cal S}$.
The resulting number of clauses is ....  , which can be roughly cut in half by  symmetry considerations.

@@

The computation of the feature pool $\cal F$ used in $T({\cal S},{\cal F})$
relies on the predicates used to encode the instances in $\Q$.
A simple grammar allows  us   to construct new predicates from them, 
and the boolean and numerical features in the pool
are defined from the resulting \emph{unary predicates}. 
The key issue is how to define new predicates from old  ones. For this 
we take a solution off-the-shelf: \emph{description logics (DL)},  called  also
\emph{concept languages},  are grammars for doing exactly this.
Concepts $C$  refer  to predicates of arity 1 and roles $R$  to predicates of arity 2.
Concept languages have been used before for  generating general policies using purely learning methods  \cite{martin:concept,fern:bias}.
For details on the syntax and semantics of description logics; see \cite{dl-handbook}.

@@

  \begin{itemize}
  \item $C \ \leftarrow\ \ C_p, C_u, C_x$: primitive, universal,  nominal c's,\footnote{Denotation of $C_x$ for parameter $x$ is $\{x\}$.}
  \item $C \ \leftarrow\  \neg C, C \sqcap C'$: negation and conjunction, 
  \item $C \ \leftarrow\  \exists R.C , \forall R.C$: concepts that use of  roles: first denotes $x$ s.t. for some  $y$,  $R(x,y) \land C(y)$.
  Second,  $x$ s.t. for all $y$, $R(x,y) \rightarrow C(y)$,
  \item $C \ \leftarrow\  R=R'$: stands for $x$ s.t. set of $y$'s  s.t. $R(x,y)$ equal to set of $y$'s s.t. $R'(x,y)$,
    \item $R \leftarrow R_p, R^{-1}, R^*$: primitive, inverse, transitive closure.
%     \item $R \leftarrow R \circ R'$, role composition: $(x,y)$ in $R \circ R'$ if $(x,z)$ in $R$ and $(z,y)$ in $R'$.
%     \item $R \leftarrow R:C$, denotes pairs $(x,y)$ in $R$ s.t. $C(y)$.
 \end{itemize}

@@

The set of all concepts and roles with complexity no larger than $k$ is denoted as ${\cal G}^k$,
with their complexity given by the minimum number of rules for generating them.
The size  of this set is reduced by incrementally pruning ``duplicate'' concepts,
those whose denotation coincides with the denotation of a another concept over all
the sampled states in ${\cal S}$.

@@

We evaluate the computational model described in steps 1--6 above over several domains.
%

% *Fill up general details in 1--6 that don't change from problem to problem
%how ${\cal S}$ obtained
The sample set $\mathcal{S}$ is obtained following one of two possible strategies.
In the \emph{exhaustive} sampling strategy,
$\mathcal{S}$ contains, for every one of the training instances $P$,
all the states generated on a breadth-first search from the initial state in $P$
that is run until $m$ expansions have been performed. If that set includes no goal state,
an optimal plan for $P$ is computed and all of the states along the plan are included in $\mathcal{S}$ as well,
as we need at least one goal state in the sample set.
%
In the \emph{random} sampling strategy, a large number $M$ of states are generated, potentially exhausting the full state space,
and then a certain number $m <\!< M$ of them is sampled uniformly at random. If no goal state is found among those $m$,
one random goal state from the larger set is drawn at random and added into the sample.

The feature complexity bound $k$ in the feature pool $\mathcal{F}$ is set to a value of $k=10$, unless no solution is found
to the theory $T({\cal S},{\cal F})$, in which case it is increased to $k=20$. 
% We have considered strategies to increase the bound adaptively
The resulting theory $T({\cal S},{\cal F})$ is solved with the Open-WBO MAXSAT solver \cite{martins2014open}.

** Details on the FOND planner, QNP translation, etc. **

The whole computational pipeline made up of steps 1--6 is processed on Intel Xeon E5-2660 CPUs,
with time and memory cutoffs of 1h and 32GB.

The number of instances $P$ is never larger than $3$, and they are selected XXX THIS NEEDS TO BE WELL JUSTIFIED...

The goal marking scheme that  yields the smaller  theory $T_G({\cal S},{\cal F})$ is obtained an optimal plan over
a few (always less than xxxx) instances $P$. ??

\begin{table*}[t]
\centering
% \addtolength{\tabcolsep}{-1pt}
{
% \scriptsize
% \footnotesize
\small
\def\arraystretch{1}%  1 is the default

\newcommand{\rowspacing[1]}{&&&&&&&&&&&&&&&\\[\dimexpr-\normalbaselineskip+2pt]}

\begin{tabular}{%
@{}l%
@{\extracolsep{1.5em}}r%@{}l%
@{\extracolsep{1.5em}}r%
@{\extracolsep{8pt}}r%
%
@{\extracolsep{1.5em}}r%
@{\extracolsep{8pt}}r%
@{\extracolsep{8pt}}r%
@{\extracolsep{8pt}}r%
%
@{\extracolsep{1.5em}}r%
@{\extracolsep{8pt}}r%
@{\extracolsep{8pt}}r%
@{\extracolsep{8pt}}r%
%
@{\extracolsep{1.5em}}r%
@{\extracolsep{8pt}}r%
@{\extracolsep{8pt}}r%
@{\extracolsep{8pt}}r%
%
@{}}

% @{\extracolsep{3pt}}

 & $I$ & $\abs{\mathcal{S}}$ & $k$ & $\abs{\mathcal{F}^k}$ & \multicolumn{2}{c}{$T$} & \multicolumn{2}{c}{$T_G$} &  $C_{\text{SAT}}$ & $\abs{F}$ & $\abs{A_F}$ & $\abs{I_F}$  & $\abs{G_F}$ & $\abs{Q^+_F}$ & $C_{\text{FOND}}$ \\ %

\rowspacing 
\cline{6-7} \cline{8-9}%
\rowspacing

 & &     &      &         & $\#V$ & $\#C$           & $\#V$ & $\#C$             &     &&&&&& \\%

\rowspacing \hline \rowspacing
\multirow{1}{*}{$\Q_{clear}$}
& 1     &     927 & 8  & 322 & 535K & 59.6M & 7.7K & 767K   &  0.01 & 3 & 2   &  - & - & - & - \\% Total complexity of F: 8

\rowspacing \hline \rowspacing
\multirow{1}{*}{$\Q_{on\text{-}2\text{-}tow}$}   
& 3 &        420  & 8 & 657 & -K & -K & 18.3K &     3.3M    &  0.02&  5 & 7   &  - & - & - & - \\% Total complexity of F: 18

\rowspacing \hline \rowspacing
\multirow{1}{*}{$\Q_{gripper}$}    
& 2 &         403 & 8 & 130 & 93K & 4.7M & 8.1K &   358K    &  0.01 & 4 & 5&  - & - & - & - \\% Total complexity of F: 10

\rowspacing \hline \rowspacing
\multirow{1}{*}{$\Q_{reward}$}
& 2 &         568 & 8 & 280 & 184K & 11.9M & 15.9K & 1.2M   &  0.01 & 2 & 2 &  - & - & - & - \\% Total complexity of F: ?
\end{tabular}

}
\caption{{
  Experimental Results on 4 different generalized planning problems. 
  $I$: number of training instances; 
  $\abs{\mathcal{S}}$: sample set size, in total number of transitions; 
  $k$: max. feature comlexity; $\abs{\mathcal{F}^k}$: size of the pool of features;
  $\#V$, $\#C$: Number of variables and clauses of the theories $T({\cal S},{\cal F})$ and $T_G({\cal S},{\cal F})$;
  $C_{\text{SAT}}$, $C_{\text{FOND}}$: computation time (sec.) of the SAT and FOND solvers, resp.;
  $\abs{F}$: number of selected features;
  $\abs{A_F}$: number of abstract actions;
  $\abs{I_F}$: ???;
  $\abs{G_F}$: ??? ;
  $\abs{Q^+_F}$: ??? .
  All figures that depend on the SAT theory are given for $T_G({\cal S},{\cal F})$.
}}
\label{tab:exp-result-data}
\end{table*}
% 2018-09-05 00:54:08 INFO     Selected (states / goals / optimal tx): 220 / 34 / 20

Table~\ref{tab:exp-result-data} summarizes the relevant data for each of the test benchmarks.

% For each problem: 1--6 details that vary from problem to problem. Follow same structure:
% $\Q$, $P$s, primitive predicates, number of atoms and clauses  of $T$ and/or $T^G$,
% SAT times,  $F$, $|A_F|$;  number of FOND atoms and actions in $Q^+_F$ (include those in $F$, $A_F$ plus translation), $I_F$ and $G_F$,
% FOND time, resulting policy.  Numbers can go all in a table.

\paragraph{Clearing a block.}
The generalized problem $\Q_{clear}$ aims at achieving the goal $clear(x)$ for any block $x$
in the Blocks world. 
We take the standard encoding with primitive predicates $on(a, b)$, $clear(b)$, $ontable(b)$, $holding(b)$ and $handempty$.

The abstract action model features one action \emph{pick-above-$x$} to pick the topmost block above $x$,
and one action \emph{place-aside} to put it away, both described in terms of features $H$, $X$ and $n(x)$, 
that denote, respectively, whether any block is being held, whether block $x$ is being held,
and the number of blocks above $x$.\footnote{
Action and features names are assigned manually by us to increase readability, but correspond faithfully
to the learnt entities.
}
%
The model is provably correct, but not complete in general, as it lacks e.g.
any action to pick block $x$.
However, it is sufficient to obtain a policy with a single loop that will 
alternatively \emph{pick-above-$x$} and \emph{place-aside} until block $x$ is clear,
which provably \emph{achieves the goal from any configuration of blocks, for any number of blocks}.

% Action 1:
%         PRE: NOT holding(a), bool[holding]
%         EFFS: NOT bool[holding]
% 
% Action 2:
%         PRE: NOT bool[holding], NOT holding(a), n(a) > 0
%         EFFS: DEC n(a), bool[holding]

% Loop:
%   Action 2 [Pick block above x]
%   If n(a) = 0 Break
%   Action 1 [Put block aside]
%

\paragraph{Stacking two blocks from different towers.}
The problem $\Q_{on\text{-}2\text{-}tow}$ is aims at achieving a goal of the form $on(x,y)$
in the Blocks world for any two blocks $x$ that are initially not in the same tower.
We take the same encoding as above.

\paragraph{Gripper.}
The problem $\Q_{gripper}$ is based on the classical Gripper problem, where a robot with two arms needs to 
move a number of blocks between two different rooms, each arm being able to carry only one ball at a time.
We use the standard encoding with primitive predicates $at\text{-}robby(l)$, $at\text{-}ball(b, l)$, $free(g)$, $carry(b, g)$ that denote,
respectively, whether the robot (the ball) is at location $l$, whether gripper $g$ is free, and whether gripper $g$ carries ball $b$,
plus type unary predicates $room$, $ball$, and $gripper$.

\paragraph{Collecting rewards in a maze.}
The objective in the problem $\Q_{reward}$ is to move in an $n \times m$ grid to pick up certain
rewards that are spread out in certain grid cells, while avoiding other cells which are considered as \emph{blocked}.
The encoding has primitive predicates $reward(c)$, $at(c)$, $blocked(c)$ and $adjacent(c_1, c_2)$ that denote
respectively the position of the rewards, of the agent, of the blocked cells, and the grid topology.
We use two single instances of the problem with size $5 \times 5$ and $4 \times 4$ and different distributions of blocks and rewards.
